Diagnostic Potential & Interactive Dynamics of Viral & Bacterial Communities in Colorectal Cancer

### 2017-06-27 doc/manuscript.Rmd
Updated for new format.

---
title: Viral and Bacterial Communities of Colorectal Cancer
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 12pt
mainfont: 'Helvetica'
---

&nbsp;

| ***Corresponding Author Information***
| Patrick D Schloss, PhD
| 1150 W Medical Center Dr. 1526 MSRB I
| Ann Arbor, Michigan 48109
| Phone: (734) 647-5801
| Email: pschloss@umich.edu

| ***Major Classification***: Biological Sciences
| ***Minor Classification***: Microbiology

***Keywords***: Colorectal Cancer, Virome, Machine Learning, Bacteriophage, Phage, Microbiome, Microbiota

@@

# Abstract

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Colorectal cancer development has been linked to differences in colonic bacterial community composition. Viruses are another important component of the colonic microbial community, however they have yet to be studied in colorectal cancer despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) in stool using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We identified no detectable difference in viral diversity (alpha or beta) between healthy, adenomatous, or cancerous colonic samples, but more sophisticated random forest models identified striking differences in the virome. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function. Our data suggest that the influential phages do not exclusively infect influential bacteria, but rather act through the community as a whole. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

# Significance Statement

Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its risk and severity have been linked to colonic bacterial community composition. Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities. We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences. The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community. These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages (viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease. Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

# Results

@@

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using OGU relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**. Because the coverage provided by the metagenomic sequencing was not as deep as the equivalent 16S rRNA gene sequencing, we attempted to compare the approaches at a common sequencing depth. This investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% across the samples **(Figure \ref{16scompare} B)**. Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

@@

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

Here, we have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

A notable finding was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences. We believe this observation speaks to the importance of sequencing coverage in microbial community studies and the advantage of the high coverage of 16S rRNA gene sequencing relative to the lower per OTU coverage possible using whole metagenomic shotgun sequencing. To demonstrate this concept, consider that six bacterial OTUs drove the performance of the 16S rRNA gene classification model and these OTUs were all sparsely present and lowly abundant. Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model. The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application. There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification [@Zeller:2014ix], but this previous approach relied on lowly abundant signatures and did not utilize OGU clustering, as done here. In that former case, the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix]. Thus, the targeted 16S rRNA gene sequencing approach, which represented only a fraction of the bacterial metagenomic sequencing depth, was more effective for detecting colorectal cancer in stool samples. Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.

@@

# Materials & Methods
This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent. A detailed description of protocols used for sample collection, processing, and analysis is provided in SI Materials and Methods. All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`. All associated source code is available at the following GitHub repository:

https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017.

# Acknowledgments
The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. GD Hannigan was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528). PD Schloss was supported by funding from the National Institutes of Health (P30DK034933). MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400).

@@

# Figures

@@

# Supplemental Information

# Supplemental Materials & Methods

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017

@@

## 16S rRNA Gene Sequence Data Acquisition & Processing
The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba]. Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC. The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do]. Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold. Abundances were normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].

@@

![*ROC curves from A) virome and B) bacterial 16S three-class random forest models tuned on mean AUC. Each curve represents the ability of the specified class to be classified against the other two classes. C) Quantification of the mean AUC variation for each model based on 10 model iterations. A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that the models are significantly different (alpha = 0.01). D) Mean decrease in accuracy when virome operational genomic units and E) bacterial 16S OTUs are removed from the respective three-class classification models. Results based on 25 iterations. F) Relative abundance of the six most important virome OVUs in the model, with the most important on the right. Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=80% }

>>>>

---
title: Diagnostic Potential & Interactive Dynamics of Viral & Bacterial Communities in Colorectal Cancer
author:
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology & Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family & Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
abstract: Colorectal cancer is a primary cause of morbidity and mortality throughout the world. Colorectal cancer development has been linked to differences in colonic bacterial community composition, however viruses have yet to be studied in colorectal cancer despite their oncogenic potential. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
font: Times
fontsize: 12pt
---

@@

## Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages (viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease. Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

@@

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using OGU relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**. Because the coverage provided by the metagenomic sequencing was not as deep as the equivalent 16S rRNA gene sequencing, we attempted to compare the approaches at a common sequencing depth. This investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% across the samples **(Figure \ref{16scompare} B)**. Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes. Thus, the targeted 16S rRNA gene sequencing approach, which represented only a fraction of the bacterial metagenomic sequencing depth, was more effective for detecting colorectal cancer in stool samples. Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.

@@

## Working Model for Virome & Cancer Progression
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

We have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

@@

## Conclusions

@@

## Acknowledgments
The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. GD Hannigan was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528). PD Schloss was supported by funding from the National Institutes of Health (P30DK034933). MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400). The authors declare no competing interests.

@@

## Figures

@@

# Supplemental Materials

# Materials & Methods

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_Science_2017

@@

## 16S rRNA Gene Sequence Data Acquisition & Processing
The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba]. Sequence (fastq) and metadata files were downloaded from:

http://www.mothur.org/MicrobiomeBiomarkerCRC

The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do]. Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold. Abundances were normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].

@@

![*ROC curves from A) virome and B) bacterial 16S three-class random forest models tuned on mean AUC. Each curve represents the ability of the specified class to be classified against the other two classes. C) Quantification of the mean AUC variation for each model based on 10 model iterations. A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that the models are significantly different (alpha = 0.01). D) Mean decrease in accuracy when virome operational genomic units and E) bacterial 16S OTUs are removed from the respective three-class classification models. Results based on 25 iterations. F) Relative abundance of the six most important virome OVUs in the model, with the most important on the right. Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=75% }
